Type-Safe Modular Parsing

### 2017-01-11 paper/sections/ObjectAlgebras.tex
Some revision for S4

The inheritance-based approach benefits us from building extensible parsers, based on an OO class hierarchy.
Nevertheless, the addition of new operations over ASTs is problematic using traditional OO classes. Often it requires modification on existing code, which breaks modularity. In this section, we will have a closer look at how Object Algebras is applied for full extensibility.

@@

Now it comes to Extension 2 in Section~\ref{subsec:parsingproblem}. In the previous code, we have pretty-printing as
an example of operations on ASTs. Suppose we also want to collect free variables on the expression language which models
literals, additions and variables. Based on the traditional OO class hierarchy, one attempt would be extending \inlinecode{Expr} with the new
operation directly:

@@

It refers to one Scala solution to the Expression Problem.\huang{what's the solution? does it have the later "critical issue"? if so, why is it a "solution"} But then programmers have to define new classes for literals, additions and variables
by extending the old ones together with \inlinecode{FreeVarsExpr}. The critical issue is that, the existing parsing code only generates objects of the old
types, again it is a modularity issue, since it is a bad idea to modify existing code again and again with extensions. Other solutions including modelling constructors as functions introduce unnecessary complexity.

@@

\subsection{Object Algebras}\label{subsec:objectalgebras}
Object Algebras captures a design pattern to address the Expression Problem nicely, achieving two dimensions of extensibility (data variants and operations) in a modular and type-safe way. The definition of data structures is separated from their behaviors, and future extensions on both dimensions no longer require existing code to be modified, supporting separate compilation.

Using Object Algebras in Scala, ASTs as recursive data structures are defined by traits, where each constructor corresponds to an abstract method inside. The example from Section~\ref{subsec:packratparsing} is used here again for illustration.
At first the language only supports literals and additions:

\lstinputlisting[linerange=5-8]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_ALG
We have two constructors \inlinecode{lit} and \inlinecode{add}, corresponding to literals and additions. The trait \inlinecode{ExprAlg} is called an \textit{Object Algebra interface},
whereas as a factory, it cannot produces objects for the expressions directly like traditional approaches, but abstract the results in its
type parameter \inlinecode{E} instead. To realize an operation on expressions, we simply instantiate the type parameter by a concrete type and
provides implementations for all cases. Below is an example of pretty-printing an expression:

@@

Futhermore, the data variants can be extended by inheriting \inlinecode{ExprAlg} and adding new cases only. Now we extend the language
with variables. A new Object Algebra interface \inlinecode{VarAlg} is defined as follows:

@@

\lstinputlisting[linerange=32-34]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_EXTPRINT
An observation is that only the new case is implemented for pretty-printing, and the others have been inherited.
It offers programmers great code reuse, in the sense that existing cases can also be
overridden in \inlinecode{VarExprPrint}.

@@

The code below results in \inlinecode{"(1 + x)"}, as the result of pretty-printing that expression.

@@

Now we observe that an ``object'' of expression actually
has the function type \inlinecode{VarExprAlg[E] =>} \inlinecode{E} for generic \inlinecode{E}. Comparing with traditional objects, one may have concern that if a parser produces function structures, it could be hard to further process them. Especially, in the design of a compiler
there will be a lot of desugarings (or transformations). However, we argue that this is not an issue, because functions can be used
in a very clever way. For example, if one wants to transform a \inlinecode{VarExprAlg} expression to an \inlinecode{ExprAlg} expression by
replacing some variables with values based on a variable environment, a transformation algebra can be implemented as follows:\haoyuan{mention in discussion that shy deals with boilerplate.}

@@

Just as shown above, \inlinecode{ExprAlg} is defined for a small language using Object Algebras. Since Object Algebras represent ``objects'' implicitly as functions like \inlinecode{ExprAlg[E] =>} \inlinecode{E} for abstract \inlinecode{E}, we first try to build the corresponding parser by type \inlinecode{ExprAlg[E] =>} \inlinecode{Parser[E]}.

>>>>

The inheritance-based approach allows building extensible parsers, based on an OO class hierarchy.  Nevertheless, the addition of new operations over ASTs is problematic using traditional OO ASTs. In this section, we 
show how to support both the forms of extensibility for ASTs (easy 
addition of language contructs, and easy addition of operations)
using Object Algebras~\cite{}.

@@

OO ASTs allow easily adding new language constructs, but it is hard 
to add new operations over the AST. The OO AST in Figure~\ref{}, 
supports a pretty-printing operation. Suppose we also want to collect free variables on the expression language which models
literals, additions and variables. Based on the traditional OO class hierarchy, one attempt would be extending \inlinecode{Expr} with the new
operation directly:

@@

\bruno{The following text is not very good, needs to be revised. I think 
we need to show the full code and where it breaks. Since I removed text 
about the Expression problem in S3, I think we need to talk more about it here.}
It refers to one Scala solution to the Expression Problem.\huang{what's the solution? does it have the later "critical issue"? if so, why is it a "solution"} But then programmers have to define new classes for literals, additions and variables
by extending the old ones together with \inlinecode{FreeVarsExpr}. The critical issue is that, the existing parsing code only generates objects of the old
types, again it is a modularity issue, since it is a bad idea to modify existing code again and again with extensions. Other solutions including modelling constructors as functions introduce unnecessary complexity.

@@

\subsection{Object Algebras}\label{subsec:objectalgebras}
Object Algebras captures a design pattern to address the Expression Problem, achieving two dimensions of extensibility (language constructs and operations) in a modular and type-safe way. The definition of data structures is separated from their behaviors, and future extensions on both dimensions no longer require existing code to be modified, supporting separate compilation.

Using Object Algebras in Scala, ASTs as recursive data structures are defined by traits, where each constructor corresponds to an abstract method inside.
Essentially, Object Algebras generalize the {\sc Abstract Factory} pattern~\cite{}, and promote the use of factory methods, instead of constructors, for instantiating objects. The example from Section~\ref{subsec:packratparsing} is used here again for illustration.
At first the language only supports literals and additions:

\lstinputlisting[linerange=5-8]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_ALG
We have two constructors \inlinecode{lit} and \inlinecode{add}, corresponding to literals and additions. The trait \inlinecode{ExprAlg} is called an \textit{Object Algebra interface}. The trait is parametrized by the type
\inlinecode{E}, which abstracts over the concrete type of the AST.

\paragraph{Adding New Operations}
To realize an operation on expressions, we simply instantiate the type parameter by a concrete type and
provides implementations for all cases. Below is an example of pretty-printing an expression:

@@

\paragraph{Adding New AST Constructs}
Futhermore, new language constructs can be added by extending \inlinecode{ExprAlg} and adding new cases only. Now we extend the language
with variables. A new Object Algebra interface \inlinecode{VarAlg} is defined as follows:

@@

\lstinputlisting[linerange=32-34]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_EXTPRINT
An observation is that only the new case is implemented for pretty-printing, and the others have been inherited.
In other words, existing code was reused and was not modified!

@@

Note how the construction of the asbtract syntax happens throught 
the use of factory methods, instead of constructors.
The pretty print the expression, resulting in \inlinecode{"(1 + x)"},
we pass an insteance of \inlinecode{VarExprPrint} to \inlinecode{makeExp}.

@@

\bruno{This is a good discussion, but it is out-of-place. We haven't 
yet talking about parsing with object algebras, so I think this discussion 
should come later.}
Now we observe that an ``object'' of expression actually
has the function type \inlinecode{VarExprAlg[E] =>} \inlinecode{E} for generic \inlinecode{E}. Comparing with traditional objects, one may have concern that if a parser produces function structures, it could be hard to further process them. Especially, in the design of a compiler
there will be a lot of desugarings (or transformations). However, we argue that this is not an issue, because functions can be used
in a very clever way. For example, if one wants to transform a \inlinecode{VarExprAlg} expression to an \inlinecode{ExprAlg} expression by
replacing some variables with values based on a variable environment, a transformation algebra can be implemented as follows:\haoyuan{mention in discussion that shy deals with boilerplate.}

@@

Parsing builds the AST as a result. When Object Algebras are used 
to build ASTs, an Object Algebra containing the constructor/factory
methods has to be passed to the parsing function.

\inlinecode{ExprAlg} is defined for a small language using Object Algebras. Since Object Algebras represent ``objects'' implicitly as functions like \inlinecode{ExprAlg[E] =>} \inlinecode{E} for abstract \inlinecode{E}, we first try to build the corresponding parser by type \inlinecode{ExprAlg[E] =>} \inlinecode{Parser[E]}.

### 2017-01-11 paper/sections/PackratParsing.tex
Some revision for S4

This section discusses the algorithmic challenges introduced by
modular parsing and argues that Packrat parser combinators 
are suitable to address those challenges. The algorithm challenges 
are important because they dictate how modular parsers should be
written. Generally speaking, the methodology we propose is that:

@@

Moreover, the underlying parsing formalism should make backtracking 
cheap, since \emph{backtracking is pervasive in modular parsing}.

>>>>

This section discusses the algorithmic challenges introduced by
modular parsing and argues that Packrat parser combinators 
are suitable to address those challenges. The algorithm challenges 
are important because they rule out various common techniques 
used by non-modular code using parsing combinators. 
To avoid pitfalls related to those algortihmic challenges,
we propose a the following methodology:

@@

Moreover, the underlying parsing formalism should make backtracking
cheap, since \emph{backtracking is pervasive in modular parsing}.
Although we chose Packrat parsing (due to its immediate availability
and support), any other parsing formalism that provides similar
features should be adequate for modular parsing.
